---
title: "Lyric Search Helpers"
author: "Tinashe M. Tapera"
date: '2018-09-21'
comments: no
showcomments: yes
showpagemeta: yes
code_folding: show
slug: lyric-search-helpers
tags:
- genuis
- dplyr
categories:
- R
- machine learning
- data science
---

Here I demonstrate the process of scraping song lyrics from the Genius API in R.

First, set up your API with the `geniusr` package. To use the Genius API, you need to authenticate your account, which you can do <a href="https://genius.com/developers">here</a> as you would with any other API service (Google if you don't understand what this means). After this, set an environment variable in your .Renvironment file as:
```
GENIUS_API_TOKEN=###your###api###token####
```
Then load `geniusr` and use the function `genius_token()` to return you genius access token to a variable:
```{r, message=FALSE, warning=FALSE}
#install.packages("tidyverse")
library(tidyverse, quietly = TRUE) #for data wrangling
#install.packages("geniusr")
library(geniusr, quietly = TRUE) #genius package

ACCESS_TOKEN = genius_token()
```

Now you can load up song lyrics and information using some of the package's handy functions and `dplyr`:
```{r}
#first, find the metadata of the artist
lupe_meta = search_artist("lupe fiasco", access_token = ACCESS_TOKEN)
lupe_meta
```

```{r}
#next, use the metadata to pull ALL of the artists songs from genius
lupe_songlist = get_artist_songs(lupe_meta$artist_id[1], access_token = ACCESS_TOKEN)
lupe_songlist
```

```{r}
#lastly, use the song name column to filter for the song you want
#then pull the song ID of that row, and use that as the id of the song
#to scrape

jump = lupe_songlist%>%
  filter(str_detect(song_name, "Jump"))%>%
  .$song_id%>%
  scrape_lyrics_id(., access_token = ACCESS_TOKEN)
jump
```

Easy peezy! Thanks to <a href=https://github.com/ewenme/geniusr>Ewen Henderson</a> for the awesome package!
